Making Sequential Programs Deterministic

### 2019-04-05 java-design.tex
Eliminate duplication

We have implemented our type system for Java, in a tool named \theDeterminismChecker.
The implementation consists of 1200 lines of Java
code, plus 1800 lines of tests (all line measurements are non-comment,
non-blank lines), a manual, examples, etc.
\TheDeterminismChecker requires Java version 8.
It is publicly available at
\myurl{URL for \theDeterminismChecker}.

@@

Our implementation treats arrays analogously to collections:
the semantics and rules are similar even though the syntax varies.
For simplicity this
paper uses the term ``collection'' and the type \<Collection> to represent
collections and arrays.
Collections include any type (including user-defined classes) that
implements the \<Iterable> or \<Iterator> interfaces; this includes all
Java collections such as \<List>s and \<Set>s.
Our implementation also supports any type that implements the \<Map>
interface.

\Cref{code-determinism} presents some of the JDK methods that we have annotated with our determinism types and give examples of
client code that would produce errors at compile time.
%\begin{figure}
%    \begin{verbatim}
%    // Annotated JDK methods
%    public class Random implements java.io.Serializable {
%        public @NonDet Random() {}
%    }
%    public class PrintStream extends FilterOutputStream 
%        implements Appendable, Closeable {
%        public void println(@Det Object x) {}
%    }
%    
%    // Client code
%    class Client {
%        void test() {
%            @Det double d = Math.random(); // Error - subtyping rules violated.
%            @NonDet double nd = Math.random(); // No error.
%            System.out.println(nd); // Error - println takes @Det arguments.
%        }
%    }
%    \end{verbatim}
%    \caption{Example: Errors detected by \theDeterminismChecker.}
%\todo{Use real examples of bugs from case studies, throughout.  Artificial
%  examples are less compelling (and lead the reader to wonder whether no
%  real example exist).  Put the first example of an error in a figure in
%  Section 1, to help with motivating the problem.}
%    \label{code-determinism}
%\end{figure}

@@

\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity(arg)> is the same as the type of
\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
  Integer>} (but see below for details about collections).
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
class ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
\todo{We do not support qualifier polymorphism on classes, only on methods.}
Java does not provide a syntax that can be used for qualifier polymorphism,
so we use a special type qualifier name, \<@PolyDet>.
A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

>>>>

We have implemented our type system for Java, in a tool named \theDeterminismChecker.
The implementation consists of 1200 lines of Java
code, plus 1800 lines of tests (all line measurements are non-comment,
non-blank lines), a manual, etc.
\TheDeterminismChecker requires Java version 8.
It is publicly available at
\myurl{URL for \theDeterminismChecker}.

@@

\TheDeterminismChecker treats arrays analogously to collections:
the semantics and rules are similar even though the syntax varies.
For simplicity this
paper uses the term ``collection'' and the type \<Collection> to represent
collections and arrays.
Collections include any type (including user-defined classes) that
implements the \<Iterable> or \<Iterator> interfaces; this includes all
Java collections such as \<List>s and \<Set>s.
Our implementation also supports any type that implements the \<Map>
interface.

\Cref{code-determinism} presents some of the JDK methods that we have annotated with our determinism types and give examples of
client code that would produce errors at compile time.

%\begin{figure}
%    \begin{verbatim}
%    // Annotated JDK methods
%    public class Random implements java.io.Serializable {
%        public @NonDet Random() {}
%    }
%    public class PrintStream extends FilterOutputStream 
%        implements Appendable, Closeable {
%        public void println(@Det Object x) {}
%    }
%    
%    // Client code
%    class Client {
%        void test() {
%            @Det double d = Math.random(); // Error - subtyping rules violated.
%            @NonDet double nd = Math.random(); // No error.
%            System.out.println(nd); // Error - println takes @Det arguments.
%        }
%    }
%    \end{verbatim}
%    \caption{Example: Errors detected by \theDeterminismChecker.}
%\todo{Use real examples of bugs from case studies, throughout.  Artificial
%  examples are less compelling (and lead the reader to wonder whether no
%  real example exist).  Put the first example of an error in a figure in
%  Section 1, to help with motivating the problem.}
%    \label{code-determinism}
%\end{figure}

@@
